A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-25 docs/paper.tex

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\%.  When viewed as a whole, the
extra memory bandwidth
and latency required to compute the simple tabulation hash are too large to
justify the dictionary speedup garnered by switching to simple tabulation hashing and linear probing.
\end{abstract}

@@

\subsection{Table Number Tuning}

@@

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the
valuable benchmarks is to see how the new dictionary performs on them. We used
PyBench benchmark suite for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

\item Simple tabulation yields slightly faster and more consistent performance: With a careful implementation simple tabulation can yield increased hash performance on linear probing work loads. Simple tabulation definitely leads to shorter chain lengths, however, those shorter chain lengths don’t always justify the performance cost.

>>>>

\begin{abstract}
Here we implement simple tabulation hashing with linear probing in Python
dictionary based on the recent theoretical result of P\v{a}tra\c{s}cu et al. Although
we achieved significant performance wins on synthetic benchmarks, on real
benchmarks our wins were only on the order of 4\% due to interpreter overhead.
Linear probing with sample tabulation hashing does improve Python performance
and should be looked at as a replacement scheme.  As a whole, the
extra memory bandwidth
and latency required to compute the simple tabulation hash negates some of the
performance improvement, but if they were able to be further mitigated, the new scheme
should provide more than a 4\% improvement, making it even more worthwhile.
\end{abstract}

@@

In order to analyze the result of this new dictionary we measured three main
metrics. First, we compared the performance of the simple tabulation hashing
with Python’s current hashing scheme on random and sequential strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
\subsection{Table Number Tuning}

@@

\subsection{Benchmarks}
Due to the heavy utilization of dictionaries in Python internals, one of the
valuable benchmarks is to see how the new dictionary performs with them them. We used
PyBench benchmark suite for this task. The results can be seen in Table
~\ref{pybenchtable}.  We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measured for the whole benchmark suite.

@@

\item Simple tabulation yields slightly faster and more consistent performance:
With a careful implementation simple tabulation can yield increased hash
performance on linear probing work loads. Simple tabulation definitely leads to
shorter chain lengths, however, those shorter chain lengths don’t always
justify the performance cost in all cases.
